### 2018-02-19 submission/manuscript.Rmd
update background

Colorectal cancer (CRC) is a growing world-wide health problem in which the microbiota has been purported to play an active role in disease pathogenesis [@siegel_cancer_2016; @flynn_metabolic_2016]. Numerous studies have shown the importance of both individual microbes [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_intestinal_2012; @kostic_fusobacterium_2013; @wu_human_2009] and the overall community [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014] in tumorgenesis using mouse models of CRC. There have also been numerous case-control studies investigating the microbiota in the formation of both adenoma and carcinoma. A recent meta-analysis investigated whether specific biomarkers could be consistently identified using multiple data sets [@shah_leveraging_2017]. This meta-analysis focused on identifying microbial signatures of tumors (biomarkers) but did so on a small total number of individuals and only investigated stool. This present meta-analysis addresses some of these major shortcomings.

Although there has been an intense focus on microbiota-based biomarker discovery for tumors, the number of candidate genera seem to be endless. Some studies point towards mouth-associated genera such as *Fusobacterium*, *Peptostreptococcus*, *Parvimonas*, and *Porphyromonas* as key enriched genera [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017]. Yet, even in these studies, mouth-associated genera are far from the only microbes identified to be associated with tumors. These other genera include, but are not limited to, *Providencia*, *Mogibacterium*, *Enterococcus*, *Escherichia/Shigella*, *Klebsiella*, and *Streptococcus* [@burns_virulence_2015; @chen_human_2012; @wang_structural_2012]. In fact, there is good *in vivo* evidence that *Escherichia/Shigella* and *Streptococcus* can be important in the pathogenesis of CRC [@ecoli_Arthur_2014; @arthur_intestinal_2012; @strep_Aymeric_2017]. Other studies have also identified *Akkermansia muciniphila* and *Bacteroides fragilis* as potential markers of CRC with good mechanistic studies for the latter [@weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014]. A recent meta-analysis confirmed the correlations of certain mouth-associated genera and *Akkermansia muciniphila* with carcinoma [@shah_leveraging_2017]. However, the sample size (n = 509) is equal to or less than some of the more recent individual studies investigating the microbiota and colorectal tumors, making it hard to know how extrapolatable these findings are. That particular meta-analysis also added more potential microbial associations to both carcinoma (*Pantoea agglomerans* *Ruminococcus*, *Lactobacillus*) and adenoma (*Prevotella*, *Methanosphaera*, *Succinovibrio*, *Haemophilus parainfluenzae*, *Ruminococcus*, *Lactobacillus*) stages of disease that need to be investigated further, since a number of these genera have been found to be enriched in controls and not disease [@baxter_microbiota-based_2016; @zeller_potential_2014; @burns_virulence_2015]. Additionally, genera like *Roseburia* have been found in some studies to be increased in tumors but in others to either be decreased or have no difference [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014].

Most of these studies have focused on individuals with carcinomas but associations with the adenoma stage of disease are not any clearer at identifying candidate genera correlated with these earlier tumors. Groups focusing on broad scale community metrics have found that measures such as richness are decreased in those with adenomas versus controls. Other studies have identified *Lactococcus*, *Pseudomonas*, *Acidovorax*, *Cloacibacterium*, *Helicobacter*, *Lactobacillus*, *Bilophila*, *Desulfovibrio*, and *Mogibacterium* to be increased in those with adenoma tumors [@sanapareddy_increased_2012; @lu_mucosal_2016; @hale_shifts_2017]. Additionally, based on these studies mentioned, there seems to be very few common genera that are associated with both adenoma and carcinoma tumors, with *Lactobacillus* being one of the few commonalities.

Targeting the identification of tumor microbial biomarkers within stool seems logical since it offers an easy and cost-effective way to stratify risk of disease. The current gold standard for diagnosis, a colonoscopy, can be time-consuming and is not without risk of complications. Although stool represents an easy and less invasive way to assess risk, it is not clear how well this sample reflects adenoma- and carcinoma- associated microbial communities. Some studies have tried to assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017]. Sampling the microbiota directly associated with colon tissue may provide clearer answers but is not without their own limitations. After the colonoscopy bowel prep the bacterial community sampled may reflect the better adhered microbiota versus the resident community. Additionally, these samples contain more host DNA, potentially limiting the types of analysis that can be done. It is well known that low biomass samples can be very difficult to work with and results can be study dependent due to the randomness of contamination [@Salter_contamination_2014].

In comparison to the previous meta-analysis, this study significantly increases the total stool samples investigated, re-examines important genera across adenoma and carcinoma across study, and examines differences and similarities between stool and tissue microbiota in the context of colorectal tumors. Importantly, this analysis and approach could provide valuable insights into the common genera that are both protective and detrimental in individuals with adenoma or carcinoma and whether broad bacterial community measurements can account for these changes that were not provided by earlier meta-analysis studies [@shah_leveraging_2017].

Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissues (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totaling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community or specific members are more important risk factors for both adenoma and carcinoma stages of disease. To accomplish this we first assessed whether bacterial diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased odds ratio (OR) for individuals to have either an adenoma or carcinoma. We then assessed what genera, if any, increase or decrease the OR of an individual to have an adenoma or carcinoma. Next, using Random Forest models, we analyzed whether the full community or only the combined top 5 increased and top 5 decreased OR genera resulted in better model classification, based on the area under the curve (AUC). Finally, we also examined at what effect and sample size the studies used were powered for and the sample size needed to get to the traditionally accepted 80% power. Our results from these analyses suggests that the bacterial community changes as disease severity worsens, that specific members are important for disease classification, and that many of the individual studies are underpowered for assessing small effect sizes.

@@

>>>>

Colorectal cancer (CRC) is a growing world-wide health problem in which the microbiota has been hypothesized to have a role in disease progression [@siegel_cancer_2016; @flynn_metabolic_2016]. Numerous studies using murine models of CRC have shown the importance of both individual microbes [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_intestinal_2012; @kostic_fusobacterium_2013; @wu_human_2009] and the overall community [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014] in tumorigenesis. Numerous case-control studies have characterized the microbiota of individuals with colonic adenomas and carcinomas in an attempt to identify biomarkers of disease progression [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017]. Because current CRC screening recommendations are poorly adhered to due to socioeconomic status, test invasiveness, and frequency of tests, development and validation of microbiome-associated biomarkers for CRC progression could further attempts to develop non-invasive diagnostics [@GimenoGarca2012].

Recently, there has been an intense focus on identifying microbiota-based biomarker yielding a seemingly endless number of candidate taxa. Some studies point towards mouth-associated genera such as Fusobacterium, Peptostreptococcus, Parvimonas, and Porphyromonas that are enriched in people with carcinomas [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017]. Other studies have identified members of *Akkermansia*, *Bacteroides*, *Enterococcus*, *Escherichia*, *Klebsiella*, *Mogibacterium*, *Streptococcus*, and *Providencia* are also associated with carciomas  [@burns_virulence_2015; @chen_human_2012; @wang_structural_2012]. Additionally, *Roseburia* has been found in some studies to be more abundant in people with tumors but in other studies it has been found to be either less abundant or no different than what is found in subjects with normal colons  [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014]. There are strong results from tissue culture and murine models that Fusobacterium nucleatum, pks-positive strains of Escherichia coli, Streptococcus gallolyticus, and an entertoxin-producing strain of Bacteroides fragilis are important in the pathogenesis of CRC [@ecoli_Arthur_2014; @arthur_intestinal_2012; @strep_Aymeric_2017; @weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014]. These results point to a causative role for the microbiota in CRC pathogenesis as well as their potential as diagnostic biomarkers.

Most studies have focused on identifying biomarkers in patients with carcinomas but there is a greater clinical need to identify biomarkers associated with adenomas. Studies focusing on broad scale community metrics have found that measures such as the total number of Operational Taxonomic Units (OTUs) are decreased in those with adenomas versus controls [@sanapareddy_increased_2012]. Other studies have identified *Acidovorax*, *Bilophila*, *Cloacibacterium*, *Desulfovibrio*, *Helicobacter*, *Lactobacillus*, *Lactococcus*, *Mogibacterium*, and *Pseudomonas* to be enriched in those with adenomas [@sanapareddy_increased_2012; @lu_mucosal_2016; @hale_shifts_2017]. There are few genera that are enriched in patients with adenoma or carcinoma tumors.

Confirming some of these previous findings, a recent meta-analysis found that 16S rRNA gene sequences from members of the *Akkermansia*, *Fusobacterium*, and *Parvimonas* were fecal biomarkers for the presence of carcinomas [@shah_leveraging_2017]. Contrary to previous studies they found sequences similar to members of *Lactobacillus* and *Ruminococcus* to be enriched in patients with adenoma or carcinoma relative to those with normal colons [@baxter_microbiota-based_2016; @burns_virulence_2015; @zeller_potential_2014]. In addition, they found 16S rRNA gene sequences from members of *Haemophilus*, *Methanosphaera*, *Prevotella*, *Succinovibrio* were enriched in patients with adenoma and *Pantoea* were enriched in patients with carcinomas. Although this meta-analysis was helpful for distilling a large number of possible biomarkers, the aggregate number of samples included in the analysis (n = 509) was smaller than several larger case-control studies that have since been published [@baxter_microbiota-based_2016; @hale_shifts_2017]

Here we provide an updated meta-analysis using 16S rRNA gene sequence data from both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) from 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2]. We expand both the breadth and scope of the previous meta-analysis to investigate whether biomarkers describing the bacterial community or specific members of the community can more accurately classify patients as having adenoma or carcinoma. Our results suggest that the bacterial community changes as disease severity worsens and that that a subset of the microbial community can be used to diagnose the presence of carcinoma.

@@

*Targeting the identification of tumor microbial biomarkers within stool seems logical since it offers an easy and cost-effective way to stratify risk of disease. The current gold standard for diagnosis, a colonoscopy, can be time-consuming and is not without risk of complications. Although stool represents an easy and less invasive way to assess risk, it is not clear how well this sample reflects adenoma- and carcinoma- associated microbial communities. Some studies have tried to assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017]. Sampling the microbiota directly associated with colon tissue may provide clearer answers but is not without their own limitations. After the colonoscopy bowel prep the bacterial community sampled may reflect the better adhered microbiota versus the resident community. Additionally, these samples contain more host DNA, potentially limiting the types of analysis that can be done. It is well known that low biomass samples can be very difficult to work with and results can be study dependent due to the randomness of contamination [@Salter_contamination_2014].*
